---
title: "Calculating z-scores"
---

## Setup

```{r}
library(here)
library(tidyverse)
library(mgcv)
library(splines)
library(RColorBrewer)
library(AUC)
library(broom)
library(ggtext)
library(ggpointdensity)
library(ggrepel)
library(ggpubr)
library(ggrastr)
```

```{r}
doc_theme <- function(...) {
  theme(axis.text = element_text(color = 'black'),
                axis.ticks = element_line(color = 'black'),
                plot.title = element_text(hjust = 0.5),
                plot.subtitle = element_text(hjust = 0.5),
                legend.key.size = unit(8, 'pt'),
                panel.background = element_blank(),
                panel.border = element_blank(),
                plot.background = element_blank(), 
                legend.background = element_blank(),
                strip.background = element_blank())
}
theme_set(theme_classic(base_size = 8, base_family = 'Arial') +
          doc_theme())
```

```{r}
lognorms <- read_csv(here('data', 'interim', 'filtered_lognorms.csv')) %>%
  select(-c(X1, pDNA_filter, `Match Bin I sum`, off_target_filter)) # X1 is old index
chip_file <- read_tsv(here('data', 'interim', 'CP0070_Chlorocebus_sabeus_remapped.chip'))
essential_genes <- read_tsv(here('data', 'raw', 'essential-genes.txt'), col_names = 'Gene')
nonessential_genes <- read_tsv(here('data', 'raw', 'non-essential-genes.txt'), col_names = 'Gene')
```

```{r}
long_lognorms <- lognorms %>%
  pivot_longer(-c(`Construct Barcode`:`Construct IDs`), 
               names_to = 'condition', values_to = 'lognorm') %>%
  filter(!grepl('Second', condition))
long_lfcs <- long_lognorms %>%
  mutate(lfc = lognorm - pDNA) %>%
  select(-c(pDNA, lognorm))
```

```{r}
# shorten labels for plotting
condition_map <- long_lfcs %>%
  select(condition) %>%
  distinct() %>%
  mutate(Cas9_old = grepl('old',condition),
         Cas9 = if_else(Cas9_old, 'Cas9-v1', 'Cas9-v2'), #Cas9-v1 refers to "old" Cas9 and Cas9-v2 refers to "new" Cas9
         low_moi = grepl('Low', condition),
         media = word(word(condition, 2, 2, sep = ';'), 1, 1), 
         density = replace_na(word(word(condition, 2, 2, sep = ';'), 2, 2), ''),
         mock = if_else(grepl('Mock', condition), 'Mock', ''),
         moi = if_else(low_moi, 'Lo-MOI', if_else(mock == 'Mock', '', 'Hi-MOI'))) %>% # Some conditions don't have a labeled MOI
  mutate(label = str_squish(paste(Cas9,media, density, mock, moi))) %>%
  select(condition, label)
```

### Heatmap of sgRNA correlations

```{r, fig.width=3, fig.height=3}
spread_lfcs <- long_lfcs %>%
  inner_join(condition_map) %>%
  select(-condition) %>%
  pivot_wider(names_from = label, values_from = lfc)
condition_cors <- spread_lfcs %>%
  select(-c(`Construct Barcode`:`Construct IDs`)) %>%
  cor()
long_cors <- condition_cors %>%
  as.data.frame() %>%
  rownames_to_column('condition.x') %>%
  pivot_longer(-condition.x, names_to = 'condition.y', values_to = 'r')
# order correlations by hierarchical clustering
cor_clust <- hclust(dist(condition_cors))
cor_order <- tibble(condition = cor_clust$labels[cor_clust$order]) %>%
  mutate(order = row_number())
long_cors <- long_cors %>%
  inner_join(cor_order, by = c('condition.x' = 'condition')) %>%
  inner_join(cor_order, by = c('condition.y' = 'condition')) %>%
  mutate(condition.x = fct_reorder(condition.x, order.x),
         condition.y = fct_reorder(condition.y, order.y))
ggplot(long_cors) +
  aes(x = condition.x, y = condition.y, fill = r) +
  geom_tile() +
  scale_fill_gradient(low = '#fff7ec', high = '#7f0000', limits = c(0,1)) +
  coord_equal() +
  xlab('condition x') +
  ylab('condition y') +
  theme(axis.text.x=element_text(angle=90, hjust=1, vjust = 0.5)) +
  labs(subtitle = 'sgRNA LFC correlations')

ggsave(here('figures', 'raw-plots', 'all_lfc_cors.svg'), width = 3, height = 3, units = 'in')
```

### ROC-AUC

Here we calculate the ROC-AUC between essentials/non-essentials for the mock infected condtions

```{r, fig.width=2.2, fig.height=2.2}
essential_chip <- chip_file %>% filter(Gene %in% essential_genes$Gene)
nonessential_chip <- chip_file %>% filter(Gene %in% nonessential_genes$Gene)
mock_roc_aucs <- long_lfcs %>%
  filter(grepl('Mock', condition)) %>%
  filter((`Construct Barcode` %in% essential_chip$`Barcode Sequence`) | # not merging to avoid duplicating guides
           (`Construct Barcode` %in% nonessential_chip$`Barcode Sequence`)) %>%
  mutate(essential = as.factor(`Construct Barcode` %in% essential_chip$`Barcode Sequence`)) %>% # otherwise nonessential
  inner_join(condition_map) %>%
  group_by(label) %>%
  mutate(pct_rank = percent_rank(-lfc)) %>% # most negative lfc = 1
  nest() %>%
  mutate(roc = map(data, function(df) {
    roc(df$pct_rank, df$essential)
    }),
    auc = map(roc, auc),
    roc_df = map(roc, tidy)) 
mock_AUCs <- mock_roc_aucs %>%
  unnest(auc) %>%
  select(-c(data, roc, roc_df))
mock_rocs <- mock_roc_aucs %>%
  unnest(roc_df) %>%
  select(-c(data, roc, auc)) %>%
  inner_join(mock_AUCs) %>%
  ungroup() %>%
  mutate(Condition = paste0(label, '\n(AUC = ', signif(auc, 2), ')'), 
         Condition = fct_reorder(Condition, -auc)) 
ggplot(mock_rocs) +
  aes(x = fpr, y = tpr, color = Condition) +
  geom_line() +
  theme(aspect.ratio = 1, legend.key.height = unit(0.6, 'cm')) +
  scale_color_brewer(palette = 'Set2') +
  scale_x_continuous(n.breaks = 3) +
  scale_y_continuous(n.breaks = 3) +
  labs(subtitle = 'sgRNA ROC') +
  ylab('tpr (essentials)') +
  xlab('fpr (non-essentials)')
ggsave(here('figures', 'raw-plots', 'roc_auc.svg'), width = 6, height = 6, units = 'cm')
```

```{r}
length(unique(essential_chip$Gene))
length(unique(nonessential_chip$Gene))

```


### Calculating Z-scores

#### Guide level

```{r}
calculate_residuals <- function(lfc_df, compare_to = 'Mock') {
  compare_cond <- lfc_df %>%
    filter(grepl(compare_to, condition)) %>%
    select(-condition) %>%
    rename(lfc_x = lfc)
  guide_residuals <- lfc_df %>%
    filter(!grepl(compare_to, condition)) %>% # don't calculate residuals for "Mock"
    inner_join(compare_cond) %>%
    group_by(condition) %>%
    mutate(residual = gam(lfc ~ ns(lfc_x, df = 4))$residual) %>%
    ungroup()
  return(guide_residuals)
}
# we calculate old and new residuals seperately, since they rely on different reference conditions
old_lfcs <- long_lfcs %>%
  filter(grepl('old', condition))
old_residuals <- calculate_residuals(old_lfcs)
new_lfcs <- long_lfcs %>%
  filter(grepl('New', condition))
new_residuals <- calculate_residuals(new_lfcs)
```

We'll use a representative condition to demonstrate the fit spline

```{r, fig.width=2.2, fig.height=2.2}
example_condition <- 'Cas9-v2 D5 5e6 Hi-MOI'
ACE2_guides <- chip_file$`Barcode Sequence`[chip_file$Gene == 'ACE2'] # not joining by chip again as to avoid duplication
get_density <- function(x, y, ...) {
  # from slowkow https://slowkow.com/notes/ggplot2-color-by-density/
  dens <- MASS::kde2d(x, y, ...)
  ix <- findInterval(x, dens$x)
  iy <- findInterval(y, dens$y)
  ii <- cbind(ix, iy)
  return(dens$z[ii])
}
example_model <- new_residuals %>%
  inner_join(condition_map) %>%
  filter(label == example_condition) %>%
  mutate(density = get_density(lfc_x, lfc, n = 100))

residual_example <- example_model %>% 
                 filter(`Construct Barcode` %in% ACE2_guides) %>%
                 top_n(1, residual) %>%
                 mutate(line_y = lfc - residual)

ggplot(example_model) +
  aes(x = lfc_x, y = lfc) +
  geom_point_rast(aes(color = density), alpha = 0.3, show.legend = F, size = 1, 
                  raster.dpi=300, pch = 16) +
  scale_color_viridis_c(option = 'D') +
  theme(aspect.ratio = 1, plot.subtitle = element_markdown()) +
  geom_point(data = example_model %>% filter(`Construct Barcode` %in% ACE2_guides), 
             aes(fill = 'ACE2'), size = 2, pch = 21) +
  scale_fill_manual(values = c('#66c2a5')) +
  geom_smooth(method = 'gam', formula = y ~ ns(x, df = 4), color = '#8da0cb', size = 1) +
  geom_segment(aes(x = lfc_x, y = lfc, xend = lfc_x, yend = line_y), 
               data = residual_example, color = 'black', 
               arrow = arrow(length = unit(4, 'pt'), ends = 'both')) +
  geom_text_repel(aes(x = lfc_x, y = (lfc+line_y)/2), data = residual_example, 
             label = 'residual', size = 2.5, nudge_x = -2.5, nudge_y = 2, family = 'Arial') +
  labs(subtitle = "sgRNA model") +
  xlab('D5 Mock (LFC)') +
  ylab(paste(word(example_condition, 2, -1), '(LFC)')) +
  guides(fill = guide_legend(title = element_blank()))
ggsave(here('figures', 'raw-plots', 'sgrna_model.pdf'), width = 5.75, 
       height = 5.75, units = 'cm', dpi = 300, device = cairo_pdf) # saving as pdf to embed rasterized plot area
```

```{r}
example_condition <- 'Cas9-v2 D5 5e6 Hi-MOI'
gene <- 'GTF2H2'
guides <- chip_file$`Barcode Sequence`[chip_file$Gene == gene] # not joining by chip again as to
residual_example <- example_model %>% 
                 filter(`Construct Barcode` %in% guides) %>%
                 top_n(1, -residual) %>%
                 mutate(line_y = lfc - residual)
ggplot(example_model) +
  aes(x = lfc_x, y = lfc) +
  geom_point_rast(aes(color = density), alpha = 0.3, show.legend = F, size = 1, 
                  raster.dpi=300, pch = 16) +
  scale_color_viridis_c(option = 'D') +
  theme(aspect.ratio = 1, plot.subtitle = element_markdown()) +
  geom_point(data = example_model %>% filter(`Construct Barcode` %in% guides), 
             aes(fill = gene), size = 2, pch = 21) +
  scale_fill_manual(values = c('#66c2a5')) +
  geom_smooth(method = 'gam', formula = y ~ ns(x, df = 4), color = '#8da0cb', size = 1) +
  geom_segment(aes(x = lfc_x, y = lfc, xend = lfc_x, yend = line_y), 
               data = residual_example, color = 'black', 
               arrow = arrow(length = unit(4, 'pt'), ends = 'both')) +
  geom_text_repel(aes(x = lfc_x, y = (lfc+line_y)/2), data = residual_example, 
             label = 'residual', size = 2.5, nudge_x = -2.5, nudge_y = 2, family = 'Arial') +
  labs(subtitle = "sgRNA model") +
  xlab('Mock (LFC)') +
  ylab(paste(example_condition, '(LFC)')) +
  guides(fill = guide_legend(title = element_blank())) +
  geom_abline()
ggsave(here('figures', 'raw-plots', paste0('SARS2', '_', example_condition, '_', gene, '_', example_condition, '_sgrna_model.pdf')), width = 5.75, 
       height = 5.75, units = 'cm', dpi = 300, device = cairo_pdf) # saving as pdf to embed rasterized
```

#### Gene level

```{r}
all_residuals <- bind_rows(new_residuals, old_residuals) %>%
  inner_join(condition_map)
pop_stats <- all_residuals %>%
  group_by(label) %>%
  summarise(pop_mean = mean(residual), pop_sd = sd(residual))
```

```{r}
gene_zscores <- all_residuals %>%
  inner_join(chip_file, by = c('Construct Barcode' = 'Barcode Sequence')) %>%
  group_by(Gene, label) %>%
  summarise(n = n(),
            mean_lfc = mean(lfc),
            mean_residual = mean(residual)) %>%
  ungroup() %>%
  inner_join(pop_stats) %>%
  mutate(z_score = (mean_residual - pop_mean)/(pop_sd/sqrt(n)),
         p_value = 2*pnorm(-abs(z_score)), 
         nlog_p_value = -log10(p_value))

```

##### Condition comparison

Based on the correlation of log2-fold changes, we average z-scores into two groups - Cas9 v1 & Cas9 v2

```{r}
v2_avg <- gene_zscores %>%
  filter(grepl('Cas9-v2', label)) %>%
  group_by(Gene) %>%
  summarise(mean_z = mean(z_score), 
            n = first(n),
            mean_lfc = mean(mean_lfc), 
            combined_pvalue = pchisq(-2 * sum(log(p_value)),df=2*length(p_value),lower=FALSE)) %>% # Fisher's method
  ungroup() %>%
  mutate(fdr = p.adjust(combined_pvalue, method = 'BH'),
         ascending_rank = rank(mean_z),
         descending_rank = rank(-mean_z)) 
v1_avg <- gene_zscores %>%
  filter(grepl('Cas9-v1', label)) %>%
  group_by(Gene) %>%
  summarise(mean_z = mean(z_score),
            ascending_rank = rank(mean_z),
            descending_rank = rank(-mean_z),
            n = first(n),
            mean_lfc = mean(mean_lfc), 
            combined_pvalue = pchisq(-2 * sum(log(p_value)),df=2*length(p_value),lower=FALSE)) %>% # Fisher's method
  ungroup() %>%
  mutate(fdr = p.adjust(combined_pvalue, method = 'BH'),
         ascending_rank = rank(mean_z),
         descending_rank = rank(-mean_z)) 
bound_mean_zs <- inner_join(v2_avg, v1_avg, by = c('Gene'), suffix = c(' Cas9v2', ' Cas9v1')) 
```

```{r, fig.width=2.2, fig.height=2.2}
control_labelled_v2 <- v2_avg %>% 
  mutate(control = grepl('NO_SITE', Gene), 
         nlogp = -log10(fdr)) %>%
         arrange(control)
top_v2 <- control_labelled_v2 %>% 
  mutate(label = if_else((rank(fdr) <= 4) | (Gene %in% c('CTSL')), Gene, NA_character_))
control_avg <- control_labelled_v2 %>%
  filter(grepl('NO_SITE', Gene)) %>%
  summarise(mean_z = mean(mean_z), nlogp = mean(nlogp))
ggplot(control_labelled_v2) +
  aes(x = mean_z, y = nlogp) +
  geom_point_rast(aes(color = control), alpha = 0.5, pch = 16, show.legend = F, 
                  size = 2, dpi = 300) +
  scale_color_manual(values = c('black', '#66c2a5')) +
  ylab('-log10(FDR)') +
  xlab('Mean z-score') +

  geom_text_repel(data = control_avg, label = 'controls', 
                  color = '#66c2a5', size = 2.5, nudge_y = 10, 
                  segment.color = '#66c2a5', family = 'Arial') + # change color
  geom_label_repel(aes(label = label), data = top_v2, 
                  size = 2.5, min.segment.length = 0.13, family = 'Arial', 
                  label.padding = 0.1, box.padding = 0.25, segment.color = 'grey50') +
  theme(aspect.ratio = 1) +
  labs(subtitle = 'Gene enrichment')
ggsave(here('figures', 'raw-plots', 'v2_gene_enrichment_ctsl.pdf'), 
       width = 5.75, height = 5.75, units = 'cm', device = cairo_pdf, dpi = 300)
```


```{r, fig.width=2, fig.height=2}
compare_conditions <- function(x, y, zs, n_genes = 5) {
  dense_zs <- zs %>%
    mutate(density := get_density(!!as.name(x), !!as.name(y), n = 100))
  ggplot(dense_zs) +
    aes(x = !!as.name(x),  y = !!as.name(y)) +
    geom_point_rast(aes(color = density), show.legend = F, 
                    alpha = 0.5, size = 1.5, pch = 16, raster.dpi = 300) +
    scale_color_viridis_c() +
    theme(aspect.ratio = 1) +
    geom_label_repel(data = dense_zs %>% mutate(label_point = 
                                            (rank(-abs(!!as.name(x))) <= n_genes) |
                                            (rank(-abs(!!as.name(y))) <= n_genes),
                                          point_label = if_else(label_point, Gene, NA_character_)),
                     aes(label = point_label), size = 2.2, min.segment.length = 0, family = 'Arial', 
                     label.padding = 0.1) +
    stat_cor(aes(label = ..r.label..), size = 2.2, family = 'Arial')
}
compare_conditions('mean_z Cas9v2', 'mean_z Cas9v1', bound_mean_zs, n_genes = 4) +
  xlab('mean z-score Cas9v2') +
  ylab('mean z-score Cas9v1')
ggsave(here('figures', 'raw-plots', 'v1_v2_comparison.pdf'), 
       width = 5.75, height = 5.75, units = 'cm', device = cairo_pdf, dpi = 300)
```

### Data outputs

```{r}
spread_zscores <- gene_zscores %>%
  pivot_wider(id_cols = c(Gene, n), names_from = label, values_from = z_score) %>%
  rename(`n_guides` = n) %>%
  inner_join(v1_avg %>%
               select(Gene, mean_z) %>%
               rename(`Cas9-v1 Avg.` = mean_z)) %>%
  inner_join(v2_avg %>%
               select(Gene, mean_z) %>%
               rename(`Cas9-v2 Avg.` = mean_z))
```

```{r}
# guide log-fold changes and residuals
write_csv(all_residuals, here('data', 'interim', 'all_guide_lfcs_residuals.csv'))
# condition mapping
write_csv(condition_map, here('data', 'interim', 'condition_mapping.csv'))
# all condition z-scores
write_csv(gene_zscores, here('data', 'interim', 'all_condition_gene_zscores.csv'))
# z-scores averaged by condition
write_csv(v2_avg, here('data', 'processed', 'Cas9_v2_avg_zscores.csv'))
write_csv(v1_avg, here('data', 'processed', 'Cas9_v1_avg_zscores.csv'))
write_csv(spread_zscores, here('data', 'processed', 'SARS-CoV-2_gw_vero_spread_zscores.csv'))
```



