---
title: "R Notebook"
---

```{r}
library(here)
library(tidyverse)
library(corrr)
library(ComplexHeatmap)
library(ggrastr)
```

```{r}
doc_theme <- function(...) {
  theme(axis.text = element_text(color = 'black'),
                axis.ticks = element_line(color = 'black'),
                plot.title = element_text(hjust = 0.5),
                plot.subtitle = element_text(hjust = 0.5),
                legend.key.size = unit(8, 'pt'),
                panel.background = element_blank(),
                panel.border = element_blank(),
                plot.background = element_blank(), 
                legend.background = element_blank(),
                strip.background = element_blank())
}
theme_set(theme_classic(base_size = 8, base_family = 'Arial') +
          doc_theme())
```

## Data 

```{r}
large_subpool_lognorms <- read_csv(here('data', 'interim', 'filtered_lognorms_cp1564.csv')) %>%
  select(-`X1`)
chip_file <- read_tsv(here('data', 'raw', 'CP1564_GRCh38_NCBI_strict_gene_20200612.chip'))
gene_pathway_mapping <- read_csv(here('data', 'interim', 'genes_pathways.csv'))
v2_avg_zscores <- read_csv(here('data', 'processed', 'Cas9_v2_avg_zscores.csv'))

```

## Growth screens

```{r}
colnames(large_subpool_lognorms) <- word(colnames(large_subpool_lognorms), 1, 1, sep='_')
col_filtered_lognorms <- large_subpool_lognorms %>%
  select(`Construct Barcode`:`EMCV #2`) # up until GFP column
```

## Calculating LFCs

We first calculate log fold changes from pDNA

```{r}
lfc_pdna <- col_filtered_lognorms %>%
  mutate(across(where(is.numeric), ~.x - pDNA)) %>%
  select(-c(pDNA))
lfc_pdna
```

We see that the two mock conditions are well correlated, so we'll combine them into one.

```{r}
get_density <- function(x, y, ...) {
  # from slowkow https://slowkow.com/notes/ggplot2-color-by-density/
  dens <- MASS::kde2d(x, y, ...)
  ix <- findInterval(x, dens$x)
  iy <- findInterval(y, dens$y)
  ii <- cbind(ix, iy)
  return(dens$z[ii])
}

dense_pdna_lfcs <- lfc_pdna %>%
  mutate(density = get_density(`Mock #1`,
                               `Mock #2`, n = 200))

ggplot(dense_pdna_lfcs) +
  aes(x = `Mock #1`, y = `Mock #2`, 
      color = density) +
  theme(aspect.ratio = 1) +
  geom_point_rast(alpha = 0.3, show.legend = F, size = 1, 
                  raster.dpi=300, pch = 16) +
  ggpubr::stat_cor(aes(label = ..r.label..), size=2.5, family='Arial') +
  scale_color_viridis_c(option = 'D') 
```

The fit spline is close enough to the abline that we'll just subtract the mock arm from all conditions

```{r}
mock_avg <- lfc_pdna %>%
    mutate(mock_avg = (`Mock #1` + 
                       `Mock #2`)/2)
ggplot(mock_avg) +
  aes(x = mock_avg, y = `SARS2 #1`) +
  geom_point() +
  geom_abline() +
  geom_smooth(method = 'gam', formula = y ~ splines::ns(x, df = 4), color = '#8da0cb', size = 1) +
  theme(aspect.ratio = 1)
```

```{r}
lfc_mock <- mock_avg %>%
  mutate(across(where(is.numeric), ~.x - mock_avg)) %>%
  select(-c(mock_avg, `Mock #1`, 
            `Mock #2`)) %>%
  rename(`HKU5-SARS-CoV-1-S #1`= `SARS1-bat #1`, 
         `HKU5-SARS-CoV-1-S #2`= `SARS1-bat #2`, 
         `SARS-CoV-2 #1` = `SARS2 #1`, 
         `SARS-CoV-2 #2` = `SARS2 #2`, 
         `VSV-SARS-CoV-2-S #1`= `rcVSV-SARS2-S #1`, 
         `VSV-SARS-CoV-2-S #2`= `rcVSV-SARS2-S #2`, 
         `MERS-CoV #1` = `MERS-WT #1`, 
         `MERS-CoV #2`= `MERS-WT #2`, 
         `MERS-CoV T1015N #1` = `MERS-T1015 #1`, 
         `MERS-CoV T1015N #2` = `MERS-T1015 #2`)
```

We see conditions grouping together as expected

```{r}
lfc_cors <- lfc_mock %>%
  select(-c(`Construct Barcode`, `Construct IDs`)) %>%
  cor()
col_fun <- circlize::colorRamp2(c(-1, 0, 1), 
                                c(scales::muted('blue'), "white", scales::muted('red')))  
pdf(here('figures', 'raw-plots', 'secondary_condition_cor.pdf'), width = 4, height = 4)
Heatmap(lfc_cors, col = col_fun, rect_gp = gpar(col = "white", lwd = 1), 
        name = 'r', column_title = 'Secondary Screens LFC Correlation',
        column_names_gp = gpar(fontsize = 6, font = 'Arial'),
        row_names_gp = gpar(fontsize = 6, font = 'Arial'),
        column_title_gp = gpar(fontsize = 7, font = 'Arial'), 
        heatmap_width = unit(7, "cm"), heatmap_height = unit(7, "cm"), 
        heatmap_legend_param = list(title_gp = gpar(fontsize = 6, font = "Arial"), 
                                    labels_gp = gpar(fontsize = 6, font = 'Arial'), 
                                    grid_height = unit(3, "mm"),
                                    grid_width = unit(3, "mm")))
dev.off()
```

## Calculating z-scores

```{r}
long_lfcs <- lfc_mock %>%
  pivot_longer(-c(`Construct Barcode`, `Construct IDs`), names_to = 'infection', 
               values_to = 'lfc') %>%
  mutate(infection = word(infection, sep = ' #'))
avg_lfcs <- long_lfcs %>%
  group_by(`Construct Barcode`, `Construct IDs`, infection) %>%
  summarise(avg_lfc = mean(lfc), .groups = 'drop')
```

We'll map controls to pseudogenes and z-score by intergenic controls

```{r}
control_genes <- chip_file %>%
  select(`Gene Symbol`) %>%
  filter(grepl('_', `Gene Symbol`)) %>%
  mutate(stub = word(`Gene Symbol`, 1, -2, sep = '_'), 
         num = as.integer(word(`Gene Symbol`, -1, -1, sep = '_')), 
         ceil_num = ceiling(num/10), 
         remapped_symbol = paste(stub, ceil_num, sep = '_'))
control_map <- deframe(control_genes %>% select(`Gene Symbol`, remapped_symbol)) # list to remap controls
```


```{r}
chip_lfcs <- avg_lfcs %>%
  inner_join(chip_file, by=c('Construct Barcode' = 'Barcode Sequence'))
ctl_remapped_chip_lfcs <- chip_lfcs %>%
  mutate(`Gene Symbol` = recode(`Gene Symbol`, !!!control_map))
intergenic_stats <- ctl_remapped_chip_lfcs %>%
  filter(grepl('ONE_INTERGENIC_SITE', `Gene Symbol`)) %>%
  group_by(infection) %>%
  summarise(ctl_mean = mean(avg_lfc), ctl_sd = sd(avg_lfc), .groups='drop')
gene_zscores <- ctl_remapped_chip_lfcs %>%
  group_by(infection, `Gene Symbol`) %>%
  summarise(avg_lfc = mean(avg_lfc), .groups='drop', 
            n = n()) %>%
  inner_join(intergenic_stats) %>%
  mutate(z_score = (avg_lfc - ctl_mean)/(ctl_sd/sqrt(n))) %>%
  select(-c(ctl_mean, ctl_sd)) %>%
  mutate(p_value = 2*pnorm(-abs(z_score))) %>%
  group_by(infection) %>%
  mutate(fdr_bh = p.adjust(p_value, method = 'BH')) %>%
  ungroup()
```

## Comparison with primary screen

```{r}
label_genes <- c('CTSL', 'TMPRSS2', 'ACE2', 'HMGB1', 'LY6E', 'SMARCA4', 'ARID1A', 'HIRA', 'CABIN1')
sars2_zs <- gene_zscores %>%
  filter(infection == 'SARS-CoV-2') %>%
  arrange(-z_score)
primary_secondary_zs <- sars2_zs %>%
  select(`Gene Symbol`, z_score) %>%
  rename(secondary_z = z_score) %>%
  inner_join(v2_avg_zscores %>%
               select(Gene, mean_z) %>%
               rename(primary_z = mean_z), by = c('Gene Symbol' = 'Gene')) %>%
  mutate(density = get_density(primary_z,
                               secondary_z, n = 100), 
         label = if_else(`Gene Symbol` %in% label_genes, `Gene Symbol`, NA_character_)) %>%
  arrange(density)
ggplot(primary_secondary_zs) +
  aes(y = primary_z, x = secondary_z) +
  theme(aspect.ratio = 1) +
  geom_point(aes(color = density), size = 0.8, pch = 16,
             show.legend = F) +
  ggpubr::stat_cor(aes(label = ..r.label..), size=2.5, family='Arial') +
  ylab('Primary screen\n(avg. residual z-score)') +
  xlab('Secondary screen\n(avg. LFC z-score)') +
  geom_hline(yintercept = 0, linetype = 'dashed') +
  geom_vline(xintercept = 0, linetype = 'dashed') +
  ggrepel::geom_label_repel(aes(label = label), size = 2.05, family = 'Arial', 
                              label.padding = 0.08, min.segment.length = 0, 
                            label.size = 0.2, box.padding = 0.1,
                            point.size = 1.5,  
                            alpha = 0.75) +

  labs(subtitle = 'SARS-CoV-2') +
  scale_color_viridis_c()
ggsave(here('figures', 'raw-plots', 'secondary_primary_cor.svg'), 
       width = 6.5, height = 6.5, units = 'cm', dpi=300)
```

## Scatter plot between conditions

```{r}
spread_zscores <- gene_zscores %>%
  pivot_wider(id_cols = `Gene Symbol`, values_from = z_score, 
              names_from = infection)
```

```{r}
compare_infections <- function(z_scores, x, y, genes = c(), n_genes = 0) {
  z_scores <- z_scores %>%
    mutate(label_point = 
             (rank(-abs(!!as.name(x))) <= n_genes) |
             (rank(-!!as.name(x)) <= n_genes) |
             (rank(-abs(!!as.name(y))) <= n_genes) |
             (rank(-!!as.name(y)) <= n_genes) |
             (rank(-!!as.name(x) - !!as.name(y) ) <= n_genes) |
             (rank(!!as.name(x) + !!as.name(y) ) <= n_genes) |
             (`Gene Symbol` %in% genes),
      label = if_else(label_point, `Gene Symbol`, NA_character_),
           density = get_density(!!as.name(x), !!as.name(y), n = 100)) %>%
    arrange(density)
  x_label <- paste(x, '(z-score)')
  y_label <- paste(y, '(z-score)')
  ggplot(z_scores) +
    aes(y = !!as.name(y), x = !!as.name(x)) +
    theme(aspect.ratio = 1) +
    geom_point(aes(color = density), size = 0.8, pch = 16, show.legend = F) +
    ylab(paste0(y, '\n(Z-scored avg. LFC)')) +
    xlab(paste0(x, '\n(Z-scored avg. LFC)')) +
    geom_hline(yintercept = 0, linetype = 'dashed') +
    geom_vline(xintercept = 0, linetype = 'dashed') +
    ggrepel::geom_label_repel(aes(label = label), size = 2.05, family = 'Arial', 
                              label.padding = 0.08, min.segment.length = 0, box.padding = 0.25, 
                              alpha = 0.75) +
    ggpubr::stat_cor(aes(label = ..r.label..), size=2.5, family='Arial') +

    scale_color_viridis_c() +
    xlab(x_label) +
    ylab(y_label) 

} 

compare_infections(spread_zscores, 'VSV-SARS-CoV-2-S', 
                   'SARS-CoV-2', genes = label_genes, n_genes = 1)
ggsave(here('figures', 'raw-plots', 'secondary_sars2_sars2vsv.svg'),
       width = 5.5, height = 5.5, units = 'cm')
```


```{r}
compare_infections(spread_zscores, 'HKU5-SARS-CoV-1-S', 
                   'SARS-CoV-2', genes = label_genes, n_genes = 1)
ggsave(here('figures', 'raw-plots', 'secondary_sars2_sars1.svg'),
       width = 5.5, height = 5.5, units = 'cm')
```

```{r}
compare_infections(spread_zscores, 'MERS-CoV', 
                   'SARS-CoV-2', genes = label_genes, n_genes = 1)
ggsave(here('figures', 'raw-plots', 'secondary_sars2_mers.svg'),
       width = 5.5, height = 5.5, units = 'cm')
```

```{r}
compare_infections(spread_zscores, 'EMCV', 
                   'SARS-CoV-2', genes = label_genes, n_genes = 1)
ggsave(here('figures', 'raw-plots', 'secondary_sars2_emcv.svg'),
       width = 5.5, height = 5.5, units = 'cm')
```

```{r}
compare_infections(spread_zscores, 'IAV-WSN', 
                   'SARS-CoV-2', genes = label_genes, n_genes = 1)
ggsave(here('figures', 'raw-plots', 'secondary_sars2_iavwsn.svg'),
       width = 5.5, height = 5.5, units = 'cm')
```

## Heatmap

```{r}
infection_order <- c('SARS-CoV-2', 'VSV-SARS-CoV-2-S', 'HKU5-SARS-CoV-1-S', 'MERS-CoV', 'MERS-CoV T1015N', 
                     'IAV-WSN', 'EMCV')
sars2_ordered_zscore <- gene_zscores %>%
  mutate(`Gene Symbol` = factor(`Gene Symbol`, levels = rev(sars2_zs$`Gene Symbol`)),
         infection = factor(infection, levels=infection_order)) 
```

```{r}
plot_term_heatmap <- function(z_scores, mapping, term) {
  term_genes <- mapping %>%
    filter(`term description` == term) %>%
    select(Gene) %>%
    unlist()
  term_zscores <- z_scores %>%
    filter(`Gene Symbol` %in% term_genes)
  ggplot(term_zscores) +
    aes(x = infection, y = `Gene Symbol`, fill = z_score) +
    geom_tile() +
    scale_fill_gradient2(midpoint = 0, 
                         high = scales::muted("red"),
                         mid = "white",
                         low = scales::muted("blue")) +
    coord_equal() +
    theme(axis.text.x=element_text(angle=90, hjust=1, vjust = 0.5)) +
    labs(subtitle = term) +
    xlab('') +
    ylab('') +
    guides(fill = guide_colorbar(title = 'Z')) 

}

plot_term_heatmap(sars2_ordered_zscore, gene_pathway_mapping, 'SWI/SNF complex')
ggsave(here('figures', 'raw-plots', 'secondary_swi_snf.svg'),
       width = 6, height = 10, units = 'cm')
```

```{r}
plot_genes_heatmap <- function(z_scores, genes, title) {
  gene_zscores <- z_scores %>%
    filter(`Gene Symbol` %in%genes)
  ggplot(gene_zscores) +
    aes(x = infection, y = `Gene Symbol`, fill = z_score) +
    geom_tile() +
    scale_fill_gradient2(midpoint = 0, 
                         high = scales::muted("red"),
                         mid = "white",
                         low = scales::muted("blue")) +
    coord_equal() +
    theme(axis.text.x=element_text(angle=90, hjust=1, vjust = 0.5)) +
    labs(subtitle = title) +
    xlab('') +
    ylab('') +
    guides(fill = guide_colorbar(title = 'Z')) 

} 

plot_genes_heatmap(sars2_ordered_zscore, c('HIRA', 'ASF1A', 'UBN1', 'CABIN1'), 
                   'HUCA complex')
ggsave(here('figures', 'raw-plots', 'secondary_huca.svg'),
       width = 6, height = 5, units = 'cm')
```

## PCA

```{r}
zscore_mat <- spread_zscores %>%
  column_to_rownames('Gene Symbol')
pca <- prcomp(zscore_mat, scale. = TRUE)
x <- 1
y <- 2
ve <- pca$sdev^2 / sum(pca$sdev^2)
PC <- paste0("PC", c(x, y))
x.column <- PC[1]
y.column <- PC[2]
loadings.column <- 'rotation'
lam <- pca$sdev[c(x, y)]
ve <- ve[c(x, y)]
labs <- paste0(PC, " (", round(ve * 100, 2), "%)")
xlab <- labs[1]
ylab <- labs[2]
pca_x <- pca$x %>%
  as.data.frame() %>%
  rownames_to_column('Gene Symbol') 
pca_label <- c(label_genes, pca_x %>% 
                 top_n(12, -!!as.name(x.column)) %>%
                 select(`Gene Symbol`) %>%
                 unlist(), 
               pca_x %>% 
                 top_n(3, !!as.name(y.column)) %>%
                 select(`Gene Symbol`) %>%
                 unlist(), 
               pca_x %>% 
                 top_n(8, !!as.name(y.column) - !!as.name(x.column)) %>%
                 select(`Gene Symbol`) %>%
                 unlist())
pca_x <- pca_x %>%
  mutate(label = if_else(`Gene Symbol` %in% pca_label, `Gene Symbol`, NA_character_))
pca_loadings <- pca[[loadings.column]] %>%
  as.data.frame() 
scaler <- min(max(abs(pca_x[[x.column]])) / max(abs(pca_loadings[[x.column]])),
              max(abs(pca_x[y.column])) / max(abs(pca_loadings[[y.column]])))
pca_loadings <- pca_loadings * scaler 
pca_loadings <- pca_loadings %>%
  rownames_to_column('condition') %>%
  as_tibble() %>%
  mutate(condition = factor(condition, levels = infection_order))
ggplot(pca_x) +
  aes(x = !!as.name(x.column), y = !!as.name(y.column)) +
  geom_point(alpha = 0.33, size = 0.8, pch = 16) +
  xlab(xlab) +
  ylab(ylab) +
  theme(aspect.ratio = ve[2]/ve[1], 
        legend.position=c(1,0), 
        legend.justification = c('right', 'bottom'), 
        legend.background = element_rect(fill = alpha('white', 0.5))) +
  geom_segment(data = pca_loadings, 
               aes(x = 0, y = 0, xend = !!as.name(x.column), yend = !!as.name(y.column), color = condition),
               arrow = grid::arrow(length = grid::unit(6, 'points'))) +
  ggrepel::geom_label_repel(aes(label = label), size = 2.1, family = 'Arial', 
                              label.padding = 0.08, min.segment.length = 0, 
                            label.size = 0.2, box.padding = 0.1, 
                            alpha = 0.8, data = pca_x %>% bind_rows(pca_loadings), point.padding = 0.05) +
  scale_color_brewer(palette = 'Set2') +
  labs(subtitle = 'Secondary Screens\n Principal Component Analysis')
ggsave(here('figures', 'raw-plots', 'secondary_pca.svg'), width = 11.75, height = 5.75, 
       units = 'cm')
```

```{r}
gene_zscores
```

## Outputs

```{r}
output_zscores <- gene_zscores %>%
  pivot_wider(id_cols = c(`Gene Symbol`, n), names_from = infection, values_from = z_score) %>%
  rename(`n_guides` = n)
```


```{r}
write_csv(output_zscores, here('data', 'processed', 'secondary_screens_z_scores_v3.csv'))
write_csv(gene_zscores, here('data', 'interim', 'secondary_screens_gene_results_v2.csv'))
```




